An Introduction to Rocker:\\ Docker Containers for R

### 2017-11-25 boettiger-eddelbuettel.Rmd
a few more tweaks

One common use case for Rocker containers is to provide a fast and
reliable mechanism to deploy a custom R environment to a remote server, such
as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
Jetstream servers \citep{jetstream}, or private or institutional server hardware.
Rocker containers are also easy to run locally on most modern laptops
using Windows, MacOS, or Linux-based operating systems. 
By sharing volumes with the local host, users can still manipulate
files with familiar, native tools while performing computation through a reproducible,
containerized environment \citep{Boettiger2015}. Being
able to test code in a predictable, pre-configured R environment on a
local machine and to then run the same code in an identical environment on a
remote server (_e.g._, for access to greater RAM, more processors, or merely
to free up the local machine from a long-running computation) is essential
for low-friction scaling of analysis.  Without such containerization,
getting code to run appropriately in a remote environment can be a major
undertaking, requiring both time and knowledge many would-be users may
not have.

@@

More details can be found in the Singularity documentation.

@@

Sharing these Dockerfiles can also facilitate the emergence 
of extensions tuned to particular communities.  For instance, the `rocker/geospatial`
image emerged from the input of a number of Rocker users all adding common
geospatial libraries and packages on top of the existing Rocker images. 
This coalescence helped create a more fine-tuned image with broad support 
for a wide range of commonly-used data formats and libraries.  Other community
images are developed and maintained independently of the Rocker project,
such as the `popgen` image of population-genetics-oriented software developed 
by the National Evolutionary Synthesis Center (NESCent). Rocker images are also 
being used as base Docker images in the NSF sponsored Whole Tale project for
reproducible computing \citep{wholetale}, and are heavily used by the R-Hub
project in automated package testing \citep{rhub}.

@@

All Rocker images are based on the Debian Linux distribution.
The Debian platform provides a small base image, the well-known `apt`
package management system, and a rich ecosystem of software libraries,
making it the base image of choice for Docker images, including many
of the "official" images maintained by Docker's own development team.
The Debian platform is also perhaps the best-supported Linux platform
within the R community, including an active `r-sig-debian` listserve.
The relatively long period between stable Debian releases
(roughly two years recently) means that software in the Debian stable 
(_e.g._, `debian:jessie`, `debian:stretch`)
releases can lag significantly behind current releases of popular
software, including R.  More recent versions of packages can be found in
the pre-release distribution, `debian:testing`, while the very latest
binary builds can be found on `debian:unstable`.   The Rocker project
can be largely divided into two stacks which address different needs,
reflected in which Debian distribution they are based on.  The first stack
is based on `debian:testing`. The second, more recently-introduced stack,
is based only on Debian stable releases.  Rocker images always point to specific
stable releases (`jessie`, `stretch`), and do not use the tag `debian:stable`,
which is a rolling tag that always points to the most
recent stable version.   The different Rocker stacks have different
aims and thus provide different images, as shown in Tables 1 & 2 below.

@@

**MRAN archives**: To facilitate installation of only contemporaneous
versions of R packages on these images, the default CRAN mirror from which
to install R packages is fixed to a snapshot of CRAN corresponding to
the last date for which that version of R was current (e.g. `3.4.2` was released
on 2017-09-28, thus `3.4.1` is pinned to the MRAN snapshot for that date). These
snapshots are provided by the MRAN archive created by Revolution
Analytics (now part of Microsoft). It archives daily snapshots
of all of CRAN from which a user can install packages with the usual
`install.packages()` function  \citep{MRAN}. Users can always override this
default by passing any current CRAN repository explicitly.  Unlike CRAN,
Bioconductor only updates its repositories through bi-annual releases
aligned to R's spring release schedule. Thus, Bioconductor packages
can be installed in the usual way using `bioclite`, which automatically
selects the Bioconductor release corresponding to the version of R in use.

**Version tags**:  The version tags are propagated throughout this
stack: _e.g._, `rocker/tidyverse:devel` will provide the currently-released
versions of the R packages in the `tidyverse` \citep{tidyverse}
installed on the nightly build of R-devel.  Developers building packages
on this stack are encouraged to tag their images accordingly as well.
Table 3 indicates which versions of R are currently available in the
stack, going back to `3.1.0`.  While older versions may be added to
the stack at a later date, we note that the MRAN snapshots began in
2014-09-17 and thus go back only to the R `3.1` era.  Each tag must
be built from a separate Dockerfile, enabling minor differences in the
build instructions to accommodate changing dependencies.  Dockerfiles for
past versions (_e.g._, prior to `3.4.2` currently) are intended to remain static
over the long term, while the tag for the current version, `latest`,
and `devel` may be tweaked to accommodate new features or dependencies.
Version tags also obey semantics so that omitting the second or third
position of the tag is identical to asking for the most recent version:
i.e. `rocker/verse:3.3` is the same as `rocker/verse:3.3.3`, and 
`rocker/verse:3` is (at the time of writing), `rocker/verse:3.4.2`.
This is accomplished using post-build hooks in Docker Hub---see examples at
<https://github.com/rocker-org/rocker-versioned/> for details.

>>>>

One common use case for Rocker containers is to provide a fast and
reliable mechanism to deploy a custom R environment to a remote server, such
as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
Jetstream servers \citep{jetstream}, or private or institutional server hardware.
Rocker containers are also easy to run locally on most modern laptops
using Windows, MacOS, or Linux-based operating systems. 
By sharing volumes with the local host, users can still manipulate
files with familiar, native tools while performing computation through a reproducible,
containerized environhment \citep{Boettiger2015}. Being
able to test code in a predictable, pre-configured R environment on a
local machine and to then run the same code in an identical environment on a
remote server (_e.g._, for access to greater RAM, more processors, or merely
to free up the local machine from a long-running computation) is essential
for low-friction scaling of analysis.  Without such containerization,
getting code to run appropriately in a remote environment can be a major
undertaking, requiring both time and knowledge many would-be users may
not have.

@@

\noindent More details can be found in the Singularity documentation.

@@

Sharing these Dockerfiles can also facilitate the emergence 
of extensions tuned to particular communities.  For instance, the `rocker/geospatial`
image emerged from the input of a number of Rocker users all adding common
geospatial libraries and packages on top of the existing Rocker images. 
This coalescence helped create a more fine-tuned image with broad support 
for a wide range of commonly-used data formats and libraries.  Other community
images are developed and maintained independently of the Rocker project,
such as the `popgen` image of population-genetics-oriented software developed 
by the National Evolutionary Synthesis Center (NESCent). Rocker images are also 
being used as base Docker images in the NSF sponsored Whole Tale project for
reproducible computing \citep{wholetale}, and are heavily used by the \CRANpkg{rhub}
project in automated package testing \citep{rhub}.

@@

All Rocker images are based on the Debian Linux distribution.
It provides a small base image, the well-known `apt`
package management system, and a rich ecosystem of software libraries,
making it the base image of choice for Docker images, including many
of the "official" images maintained by Docker's own development team.
The Debian platform is also perhaps the best-supported Linux platform
within the R community, including an active `r-sig-debian` listserve.
The relatively long period between stable Debian releases
(roughly two years recently) means that software in the Debian stable 
(_e.g._, `debian:jessie`, `debian:stretch`)
releases can lag significantly behind current releases of popular
software, including R.  More recent versions of packages can be found in
the pre-release distribution, `debian:testing`, while the very latest
binary builds can be found on `debian:unstable`.   The Rocker project
can be largely divided into two stacks which address different needs,
reflected in which Debian distribution they are based on.  The first stack
is based on `debian:testing`. The second, more recently-introduced stack,
is based only on Debian stable releases.  Rocker images always point to specific
stable releases (`jessie`, `stretch`), and do not use the tag `debian:stable`,
which is a rolling tag that always points to the most
recent stable version.   The different Rocker stacks have different
aims and thus provide different images, as shown in Tables 1 & 2 below.

@@

**MRAN archives**: To facilitate installation of only contemporaneous
versions of R packages on these images, the default CRAN mirror from which
to install R packages is fixed to a snapshot of CRAN corresponding to
the last date for which that version of R was current (_e.g._, `3.4.2` was released
on 2017-09-28, thus `3.4.1` is pinned to the MRAN snapshot for that date). These
snapshots are provided by the MRAN archive created by Revolution
Analytics (now part of Microsoft). It archives daily snapshots
of all of CRAN from which a user can install packages with the usual
`install.packages()` function  \citep{MRAN}. Users can always override this
default by passing any current CRAN repository explicitly.  Unlike CRAN,
Bioconductor only updates its repositories through bi-annual releases
aligned to R's spring release schedule. Thus, Bioconductor packages
can be installed in the usual way using `bioclite`, which automatically
selects the Bioconductor release corresponding to the version of R in use.

**Version tags**:  The version tags are propagated throughout this
stack: _e.g._, `rocker/tidyverse:devel` will provide the currently-released
versions of the R packages in the \CRANpkg{tidyverse} \citep{tidyverse}
installed on the nightly build of R-devel.  Developers building packages
on this stack are encouraged to tag their images accordingly as well.
Table 3 indicates which versions of R are currently available in the
stack, going back to `3.1.0`.  While older versions may be added to
the stack at a later date, we note that the MRAN snapshots began in
2014-09-17 and thus go back only to the R `3.1` era.  Each tag must
be built from a separate Dockerfile, enabling minor differences in the
build instructions to accommodate changing dependencies.  Dockerfiles for
past versions (_e.g._, prior to `3.4.2` currently) are intended to remain static
over the long term, while the tag for the current version, `latest`,
and `devel` may be tweaked to accommodate new features or dependencies.
Version tags also obey semantics so that omitting the second or third
position of the tag is identical to asking for the most recent version:
_i.e._, `rocker/verse:3.3` is the same as `rocker/verse:3.3.3`, and 
`rocker/verse:3` is (at the time of writing), `rocker/verse:3.4.2`.
This is accomplished using post-build hooks in Docker Hub---see examples at
<https://github.com/rocker-org/rocker-versioned/> for details.

### 2017-11-25 boettiger-eddelbuettel.tex
a few more tweaks

One common use case for Rocker containers is to provide a fast and
reliable mechanism to deploy a custom R environment to a remote server,
such as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean,
NSF's Jetstream servers \citep{jetstream}, or private or institutional
server hardware. Rocker containers are also easy to run locally on most
modern laptops using Windows, MacOS, or Linux-based operating systems.
By sharing volumes with the local host, users can still manipulate files
with familiar, native tools while performing computation through a
reproducible, containerized environment \citep{Boettiger2015}. Being
able to test code in a predictable, pre-configured R environment on a
local machine and to then run the same code in an identical environment
on a remote server (\emph{e.g.}, for access to greater RAM, more
processors, or merely to free up the local machine from a long-running
computation) is essential for low-friction scaling of analysis. Without
such containerization, getting code to run appropriately in a remote
environment can be a major undertaking, requiring both time and
knowledge many would-be users may not have.

@@

To make extensions transparent and persistent, Rocker images can be
extended by any user by writing their own Dockerfiles based on an
appropriate Rocker image. The Dockerfiles in the Rocker stack should
themselves provide a simple example of this, (as described in the
following section).\\
A user begins by selecting an appropriate base image for their needs: if
the RStudio\textsuperscript{\textregistered} interface is desired, a
user might start with \texttt{FROM\ rocker/rstudio}; an image for
testing an R package with compiled code might use
\texttt{FROM\ rocker/r-devel-san}, and an image for reproducing a data
analysis will probably select a stable version tag in addition to an
appropriate base library, \emph{e.g.},:
\texttt{FROM\ rocker/tidyverse:3.4.1} Users can easily add additional
software to any running Rocker image using the standard R and Debian
mechanisms. Details on how to extend Rocker images can be found at
\url{https://rocker-project.org}.

Sharing these Dockerfiles can also facilitate the emergence of
extensions tuned to particular communities. For instance, the
\texttt{rocker/geospatial} image emerged from the input of a number of
Rocker users all adding common geospatial libraries and packages on top
of the existing Rocker images. This coalescence helped create a more
fine-tuned image with broad support for a wide range of commonly-used
data formats and libraries. Other community images are developed and
maintained independently of the Rocker project, such as the
\texttt{popgen} image of population-genetics-oriented software developed
by the National Evolutionary Synthesis Center (NESCent). Rocker images
are also being used as base Docker images in the NSF sponsored Whole
Tale project for reproducible computing \citep{wholetale}, and are
heavily used by the R-Hub project in automated package testing
\citep{rhub}.

@@

All Rocker images are based on the Debian Linux distribution. The Debian
platform provides a small base image, the well-known \texttt{apt}
package management system, and a rich ecosystem of software libraries,
making it the base image of choice for Docker images, including many of
the ``official'' images maintained by Docker's own development team. The
Debian platform is also perhaps the best-supported Linux platform within
the R community, including an active \texttt{r-sig-debian} listserve.
The relatively long period between stable Debian releases (roughly two
years recently) means that software in the Debian stable (\emph{e.g.},
\texttt{debian:jessie}, \texttt{debian:stretch}) releases can lag
significantly behind current releases of popular software, including R.
More recent versions of packages can be found in the pre-release
distribution, \texttt{debian:testing}, while the very latest binary
builds can be found on \texttt{debian:unstable}. The Rocker project can
be largely divided into two stacks which address different needs,
reflected in which Debian distribution they are based on. The first
stack is based on \texttt{debian:testing}. The second, more
recently-introduced stack, is based only on Debian stable releases.
Rocker images always point to specific stable releases (\texttt{jessie},
\texttt{stretch}), and do not use the tag \texttt{debian:stable}, which
is a rolling tag that always points to the most recent stable version.
The different Rocker stacks have different aims and thus provide
different images, as shown in Tables 1 \& 2 below.

@@

As the names \texttt{testing} and \texttt{unstable} imply, particular
versions of package can change as packages move from \texttt{unstable}
into \texttt{testing}. New versions are sent to \texttt{unstable} during
the normal course of Debian development. This can occasionally break an
previously-working installation command in a Dockerfile until the
maintainer redirects the package manager to install a package from the
\texttt{unstable} sources that could previously be installed from
\texttt{testing}, or vice versa (using the \texttt{-t} option in
\texttt{apt}). That said, packages only migrate from \texttt{unstable}
to \texttt{testing} after a period of several days---and if the
migration and installation of the particular version is free of
interactions with other packages in their dependency graph. That way,
\texttt{unstable} serves as validation lab which leaves \texttt{testing}
reasonably stable yet current.

@@

\textbf{MRAN archives}: To facilitate installation of only
contemporaneous versions of R packages on these images, the default CRAN
mirror from which to install R packages is fixed to a snapshot of CRAN
corresponding to the last date for which that version of R was current
(e.g. \texttt{3.4.2} was released on 2017-09-28, thus \texttt{3.4.1} is
pinned to the MRAN snapshot for that date). These snapshots are provided
by the MRAN archive created by Revolution Analytics (now part of
Microsoft). It archives daily snapshots of all of CRAN from which a user
can install packages with the usual \texttt{install.packages()} function
\citep{MRAN}. Users can always override this default by passing any
current CRAN repository explicitly. Unlike CRAN, Bioconductor only
updates its repositories through bi-annual releases aligned to R's
spring release schedule. Thus, Bioconductor packages can be installed in
the usual way using \texttt{bioclite}, which automatically selects the
Bioconductor release corresponding to the version of R in use.

\textbf{Version tags}: The version tags are propagated throughout this
stack: \emph{e.g.}, \texttt{rocker/tidyverse:devel} will provide the
currently-released versions of the R packages in the \texttt{tidyverse}
\citep{tidyverse} installed on the nightly build of R-devel. Developers
building packages on this stack are encouraged to tag their images
accordingly as well. Table 3 indicates which versions of R are currently
available in the stack, going back to \texttt{3.1.0}. While older
versions may be added to the stack at a later date, we note that the
MRAN snapshots began in 2014-09-17 and thus go back only to the R
\texttt{3.1} era. Each tag must be built from a separate Dockerfile,
enabling minor differences in the build instructions to accommodate
changing dependencies. Dockerfiles for past versions (\emph{e.g.}, prior
to \texttt{3.4.2} currently) are intended to remain static over the long
term, while the tag for the current version, \texttt{latest}, and
\texttt{devel} may be tweaked to accommodate new features or
dependencies. Version tags also obey semantics so that omitting the
second or third position of the tag is identical to asking for the most
recent version: i.e. \texttt{rocker/verse:3.3} is the same as
\texttt{rocker/verse:3.3.3}, and \texttt{rocker/verse:3} is (at the time
of writing), \texttt{rocker/verse:3.4.2}. This is accomplished using
post-build hooks in Docker Hub---see examples at
\url{https://github.com/rocker-org/rocker-versioned/} for details.

>>>>

One common use case for Rocker containers is to provide a fast and
reliable mechanism to deploy a custom R environment to a remote server,
such as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean,
NSF's Jetstream servers \citep{jetstream}, or private or institutional
server hardware. Rocker containers are also easy to run locally on most
modern laptops using Windows, MacOS, or Linux-based operating systems.
By sharing volumes with the local host, users can still manipulate files
with familiar, native tools while performing computation through a
reproducible, containerized environhment \citep{Boettiger2015}. Being
able to test code in a predictable, pre-configured R environment on a
local machine and to then run the same code in an identical environment
on a remote server (\emph{e.g.}, for access to greater RAM, more
processors, or merely to free up the local machine from a long-running
computation) is essential for low-friction scaling of analysis. Without
such containerization, getting code to run appropriately in a remote
environment can be a major undertaking, requiring both time and
knowledge many would-be users may not have.

@@

To make extensions transparent and persistent, Rocker images can be
extended by any user by writing their own Dockerfiles based on an
appropriate Rocker image. The Dockerfiles in the Rocker stack should
themselves provide a simple example of this, (as described in the
following section). A user begins by selecting an appropriate base image
for their needs: if the RStudio\textsuperscript{\textregistered}
interface is desired, a user might start with
\texttt{FROM\ rocker/rstudio}; an image for testing an R package with
compiled code might use \texttt{FROM\ rocker/r-devel-san}, and an image
for reproducing a data analysis will probably select a stable version
tag in addition to an appropriate base library, \emph{e.g.},:
\texttt{FROM\ rocker/tidyverse:3.4.1}. Users can easily add additional
software to any running Rocker image using the standard R and Debian
mechanisms. Details on how to extend Rocker images can be found at
\url{https://rocker-project.org}.

Sharing these Dockerfiles can also facilitate the emergence of
extensions tuned to particular communities. For instance, the
\texttt{rocker/geospatial} image emerged from the input of a number of
Rocker users all adding common geospatial libraries and packages on top
of the existing Rocker images. This coalescence helped create a more
fine-tuned image with broad support for a wide range of commonly-used
data formats and libraries. Other community images are developed and
maintained independently of the Rocker project, such as the
\texttt{popgen} image of population-genetics-oriented software developed
by the National Evolutionary Synthesis Center (NESCent). Rocker images
are also being used as base Docker images in the NSF sponsored Whole
Tale project for reproducible computing \citep{wholetale}, and are
heavily used by the \CRANpkg{rhub} project in automated package testing
\citep{rhub}.

@@

All Rocker images are based on the Debian Linux distribution. It
provides a small base image, the well-known \texttt{apt} package
management system, and a rich ecosystem of software libraries, making it
the base image of choice for Docker images, including many of the
``official'' images maintained by Docker's own development team. The
Debian platform is also perhaps the best-supported Linux platform within
the R community, including an active \texttt{r-sig-debian} listserve.
The relatively long period between stable Debian releases (roughly two
years recently) means that software in the Debian stable (\emph{e.g.},
\texttt{debian:jessie}, \texttt{debian:stretch}) releases can lag
significantly behind current releases of popular software, including R.
More recent versions of packages can be found in the pre-release
distribution, \texttt{debian:testing}, while the very latest binary
builds can be found on \texttt{debian:unstable}. The Rocker project can
be largely divided into two stacks which address different needs,
reflected in which Debian distribution they are based on. The first
stack is based on \texttt{debian:testing}. The second, more
recently-introduced stack, is based only on Debian stable releases.
Rocker images always point to specific stable releases (\texttt{jessie},
\texttt{stretch}), and do not use the tag \texttt{debian:stable}, which
is a rolling tag that always points to the most recent stable version.
The different Rocker stacks have different aims and thus provide
different images, as shown in Tables 1 \& 2 below.

@@

As the names \texttt{testing} and \texttt{unstable} imply, particular
versions of package can change as packages move from \texttt{unstable}
into \texttt{testing}. New versions are sent to \texttt{unstable} during
the normal course of Debian development. This can occasionally break a
previously-working installation command in a Dockerfile until the
maintainer redirects the package manager to install a package from the
\texttt{unstable} sources that could previously be installed from
\texttt{testing}, or vice versa (using the \texttt{-t} option in
\texttt{apt}). That said, packages only migrate from \texttt{unstable}
to \texttt{testing} after a period of several days---and if the
migration and installation of the particular version is free of
interactions with other packages in their dependency graph. That way,
\texttt{unstable} serves as validation lab which leaves \texttt{testing}
reasonably stable yet current.

@@

\textbf{MRAN archives}: To facilitate installation of only
contemporaneous versions of R packages on these images, the default CRAN
mirror from which to install R packages is fixed to a snapshot of CRAN
corresponding to the last date for which that version of R was current
(\emph{e.g.}, \texttt{3.4.2} was released on 2017-09-28, thus
\texttt{3.4.1} is pinned to the MRAN snapshot for that date). These
snapshots are provided by the MRAN archive created by Revolution
Analytics (now part of Microsoft). It archives daily snapshots of all of
CRAN from which a user can install packages with the usual
\texttt{install.packages()} function \citep{MRAN}. Users can always
override this default by passing any current CRAN repository explicitly.
Unlike CRAN, Bioconductor only updates its repositories through
bi-annual releases aligned to R's spring release schedule. Thus,
Bioconductor packages can be installed in the usual way using
\texttt{bioclite}, which automatically selects the Bioconductor release
corresponding to the version of R in use.

\textbf{Version tags}: The version tags are propagated throughout this
stack: \emph{e.g.}, \texttt{rocker/tidyverse:devel} will provide the
currently-released versions of the R packages in the \CRANpkg{tidyverse}
\citep{tidyverse} installed on the nightly build of R-devel. Developers
building packages on this stack are encouraged to tag their images
accordingly as well. Table 3 indicates which versions of R are currently
available in the stack, going back to \texttt{3.1.0}. While older
versions may be added to the stack at a later date, we note that the
MRAN snapshots began in 2014-09-17 and thus go back only to the R
\texttt{3.1} era. Each tag must be built from a separate Dockerfile,
enabling minor differences in the build instructions to accommodate
changing dependencies. Dockerfiles for past versions (\emph{e.g.}, prior
to \texttt{3.4.2} currently) are intended to remain static over the long
term, while the tag for the current version, \texttt{latest}, and
\texttt{devel} may be tweaked to accommodate new features or
dependencies. Version tags also obey semantics so that omitting the
second or third position of the tag is identical to asking for the most
recent version: \emph{i.e.}, \texttt{rocker/verse:3.3} is the same as
\texttt{rocker/verse:3.3.3}, and \texttt{rocker/verse:3} is (at the time
of writing), \texttt{rocker/verse:3.4.2}. This is accomplished using
post-build hooks in Docker Hub---see examples at
\url{https://github.com/rocker-org/rocker-versioned/} for details.
